Bayesian Restricted Likelihood Methods: Conditioning on Insufficient Statistics in Bayesian Regression\thanksref{T1}

### 2021-02-01 revision_2/revision_Dec2020 Dec 30.tex
git cleanup

% commands for for editing
\usepackage{color}
\usepackage{ulem} % \sout
\newcommand{\red}[1]{{\color{red}#1}}
\newcommand{\blue}[1]{{\color{blue}#1}}
\newcommand{\brown}[1]{{\color{brown}#1}}
\newcommand{\green}[1]{{\color{green}#1}}
\newcommand{\magenta}[1]{{\color{magenta}#1}}
\newcommand{\response}[1]{{\color{blue}#1}}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\graphicspath{{figures/}{figs/}}

@@

Imperfection in the likelihood has also been widely discussed in the classical robustness literature. \cite{hampel1971}, writing on the motivation for studies of robustness, provides a concise description of three mismatches between data and the model that purportedly gives rise to the data:  ``(i) rounding of the observations; (ii) the occurrence of gross errors; (iii) the model itself may only be an approximation to the underlying chance mechanism''.  In a Bayesian setting, the first is easily handled with MCMC methods through the introduction of a latent, unrounded variable into the model.  We do not consider it here.  The second and third are duals.  Misspecification of the model (iii) will often make observations appear to be outliers (ii).  The literature on robust methods is replete with examples described in terms of ``outliers'' where the central problem is model misspecification.  In the sequel, we follow the tradition of referring to cases that are discordant with the stated model as ``outliers'', whether this discordance is due to gross error or a consequence of model misspecification.

In practice, the imperfections in a proposed likelihood often show themselves through the presence of outliers -- whether due to local misspecification of the model or due to gross error. There are three main solutions to Bayesian outlier-handling.  The first is to replace the basic sampling density with a mixture model which includes one component for the ``good'' data and a second component for the ``bad'' data.  With this approach, the good component of the sampling density is used for prediction of future good data.  The second approach replaces the
basic sampling density with a thick-tailed density in an attempt to discount outliers, yielding techniques that often provide solid estimates of the center of the distribution but do not easily translate to predictive densities for further good data.  The third approach fits a flexible (typically nonparametric) model to  the data, producing a Bayesian version of a density estimate for both good and bad data.  In recent development, inference is made through the use of robust inference functions \citep{lee2014}.

@@

Restricted likelihoods have also been used as practical approximations to a full likelihood. For example, \cite{pratt1965} appeals to heuristic arguments regarding approximate sufficiency to justify the use of the restricted likelihood of the sample mean and standard deviation. Approximate sufficiency is also appealed to in the use of Approximate Bayesian Computation (ABC), which is related to our method.  
ABC is a collection of posterior approximation methods which has recently experienced success in applications to epidemiology, genetics, and quality control \citep[see, for example,][]{tavare1997, pritchard1999, beaumont2002, marjoram2003, fearnhead2012, drovandi2015}. Interest typically lies in the full data posterior and ABC is used for computational convenience as an approximation.  Consequently, effort is made to choose an approximately sufficient $T(\by)$ and update to the ABC posterior by using the likelihood $L(\bth| \mathcal{B}(\by))$, where $\mathcal{B}(\by)=\{\by^{*}|\rho(T(\by),T(\by^{*})) \leq \epsilon\}$, $\rho$ is a metric, and $\epsilon$ is a tolerance level. This is the likelihood conditioned on the collection of data sets that result in a $T(\cdot)$ within $\epsilon$ of the observed $T(\by)$. %Hence, we can see that, as in our method, conditioning is done on a summary of the complete data. 
With an approximately sufficient $T(\cdot)$ and a small enough $\epsilon$, heuristically  $L(\bth|\mathcal{B}(\by))\approx L(\bth|T(\by))\approx L(\bth|\by)$. Consequently, the ABC posterior approximates the full data posterior and efforts have been made to formalize what is meant by approximate sufficiency \citep[e.g.,][]{joyce2008}.

Our method can be viewed as ABC with $\epsilon = 0$ and it is natural to compare it to ABC. This paper develops sampling methods for fitting Bayesian linear models conditioning exactly on a set of summary statistics ($\epsilon = 0$), even when the statistics follow a continuous distribution. Traditional ABC sampling methods are flexible and will, in general, apply to a broader class of models. The basic sampling method for ABC is the rejection sampling algorithm \citep{pritchard1999} which proposes a sample $\bth^{*}$ from the prior, then new data $\by^{*}$ from the data-model given $\bth^{*}$. The value $\bth^{*}$ is accepted as a draw from the ABC posterior if $\rho(T(\by),T(\by^{*})) \leq \epsilon$. Acceptance rates of this algorithm can be intolerably low and several extensions have been proposed to improve efficiency \citep[see, for example,][]{beaumont2009, turner2012}. The inefficiency of ABC algorithms is especially problematic in high-dimensional settings since generating high-dimensional statistics that are close to the observed values is difficult. Recently, \cite{turner2014} developed the Gibbs ABC method which improves efficiency in the hierarchical setting by making use of conditional independence of the model to make accept/reject decisions at the individual group-level, effectively reducing the dimension of the problem to the number of parameters within each group. We revisit this approach in our comparisons to ABC in Section \ref{hierRegNW}, finding that, for a modest increase in comptational cost, we obtain an algorithm with better convergence and mixing properties.  We also retain the desired posterior distribution -- the posterior, having conditioned exactly on the summary statistics.

@@

\subsection{Illustrative Examples}
\label{illustrations}
Before discussing computational details, the method is applied to two simple examples on well known data sets to demonstrate its effectiveness in situations where outliers are a major concern. The full model in each case fits into the Bayesian linear regression framework discussed in Section \ref{BayesLinMod}. The first is an example (so far as we know) of gross error; the second is an example of model misspecification for a subset of the observations.  The first example is an analysis of Simon Newcomb's 66 measurements of the passage time of light \citep{stigler1977}; two of which are significant outliers in the lower tail. The full model is a standard location-scale Bayesian model also used in \cite{lee2014}:
The full model in each case fits into the Bayesian linear regression framework discussed in Section \ref{BayesLinMod}. The first example is an analysis of Simon Newcomb's 66 measurements of the passage time of light \citep{stigler1977}; two of which are significant outliers in the lower tail. The full model is a standard location-scale Bayesian model also used in \cite{lee2014}:\begin{equation}
\beta\sim N(23.6, 2.04^{2}),\  \sigma^{2}\sim IG(5, 10), \ y_{i}\iid N (\beta, \sigma^{2}), i=1,2,\dots, n=66,
\end{equation}
where $y_{i}$ denotes the $i^{th}$ (recorded) measurement of the passage time of light. $\beta$ is interpreted as the passage time of light with the deviations $y_i - \beta$ representing measurement error.
Four versions of the restricted likelihood are fit with conditioning statistics: 1) Huber's M-estimator for location with Huber's `proposal 2'  for scale 2)  Tukey's M-estimator for location with Huber's `proposal 2'  for scale 3) LMS (least median squares) for location with associated estimator of scale and 4) LTS (least trimmed squares)  for location with associated estimator of scale. Details of these estimators can be found in many places, including \citep{huber2009}. We return to the two M-estimators throughout this paper as we have found them to offer good default choices for practitioners dealing with outliers. A short review of these estimators is provided in the Supplementary Material.  The tuning parameters for the M-estimators are chosen to achieve $95\%$ efficiency under normality \citep{huber2009} and, for comparability, roughly $5\%$ of the residuals are trimmed for LTS.  Two additional approaches to outlier handling are considered: 1) the normal distribution is replaced with a t-distribution and, 2) the normal distribution is replaced with a mixture of two normals. The t-model assumes $y_{i}\iid t_{\nu} (\beta, \sigma^{2})$ with $\nu=5$. The prior on $\sigma^{2}$ is $IG(5, \frac{\nu-2}{\nu}10)$ and ensures that the prior on the variance is the same as the other models. The mixture takes the form: $y_{i}\iid pN (\beta, \sigma^{2}) + (1-p)N(\beta, 10\sigma^{2})$ with the prior $p \sim \text{beta}(20,1)$ on the probability of belonging to the `good' component.

The posterior of $\beta$ under each model appears in Figure \ref{fig:newcomb_post}.  The posteriors group into two batches.  The normal model and restricted likelihood with LMS do not discount the outliers and have posteriors centered at low values of $\beta$.  These posteriors are also quite diffuse.  In contrast, the t-model, mixture model, and the other restricted likelihood methods discount the outliers and have posteriors centered at higher values.  There is modest variation among these centers.  Posteriors in this second group have less dispersion than those in the first group.   
\begin{figure}[t]
\centering
{\includegraphics[width = 4in]{figs/speed_of_light_beta.png}}
{\includegraphics[width = 4in]{figs/speed_of_light_predictive.png}}
\caption{Results from the analysis of the speed of light data. Top: Posterior distributions of $\beta$ under each model. Bottom: Log posterior predictive distributions under each model. The differences in the tails are emphasized in the bottom plot. The horizontal axis is strategically labeled to help compare the centers of the distributions in each of the plots.}
\label{fig:newcomb_post}
\end{figure}
The pattern for predictive distributions differs (see bottom plot in Figure \ref{fig:newcomb_post}).  The normal and t-models have widely dispersed predictive distributions.  The other predictive distributions show much greater concentration.  The restricted likelihood fits based on M-estimators (Tukey's and Huber's) are centered appropriately and are concentrated. The restricted likelihood based on LTS and the mixture model results are also centered appropriately, but comparatively less concentrated. The LMS predictive is concentrated, but it is poorly centered.

@@

The data and  $95\%$ credible bands for the posterior predictive distribution under each model are displayed in Figure \ref{fig:calls_predictive}. The normal model fit to all cases results in a very wide posterior predictive distribution due to an inflated estimate of the variance. The t-model provides a similar predictive distribution.  The pocket of outliers from 1963 to 1970 overwhelms the natural robustness of the model and leads to wide prediction bands.  The outliers, falling toward the end of the time period, lead to a relatively high slope for the regression.  In contrast, the normal theory model fit to only the good data results in a smaller slope and narrower prediction bands.  The predictive distribution under the restricted likelihood approach is much more precise and is close to that of the normal theory fit to the non-outlying cases. The two component mixture model provides similar results, where the predictive distribution is formulated using only the good component. For these data, the large outliers are easily identified as following a distinct regression, leaving the primary component of the mixture for non-outlying data.  In a more complex situation where the outlier generating mechanism is transient (i.e., ever changing and more complex than for these data), modeling the outliers is more difficult. As in classical robust estimation, the restricted likelihood approach avoids explicitly modeling the outliers. 
\begin{figure}[t]
\centering
{\includegraphics[width = 4in]{figs/calls_predictive.png}}
\caption{Pointwise posterior predictive intervals of log(calls) under the normal theory model fit to the non-outliers, the restricted likelihood model with Tukey's M-estimator for the slope and intercept with Huber's `proposal 2'  for scale, and a heavy-tailed t-distribution model. The first three data points were used to specify the prior with each model using the remaining 21 for fitting. The normal theory model was also fit after removing observations 14-20 (years 1963 - 1970).}
\label{fig:calls_predictive}
\end{figure}

@@

Given the one-to-one and onto mapping $h: \mathbb{S} \rightarrow \mathcal{A}$, the general proposal strategy is summarized as follows:
\begin{enumerate}
\item Sample $\bz^*$ from a distribution with known density whose support is the entirety of $\mathbb{S}$.
\item Set $\by = h(\bz^*)$ and calculate the Jacobian of this transformation in two steps.
\begin{enumerate}
\item Scale from $\mathbb{S}$ to the set $\Pi(\mathcal{A}):= \{\bz\in \mathbb{R}^n |\ \exists\ \by\in \mathcal{A}\ s.t.\ \bz=Q \by \}$ with $Q = I - XX^{\top}$. \footnote{We have used condition \ref{fullRank} to assume  without loss of generality  that the columns of $X$ form an orthonormal basis for $\mc{C}(X)$ (i.e., $X^\top X=I$).} $\Pi(\mathcal{A})$ is the projection of $\mathcal{A}$ onto $\mathcal{C}^{\perp}(X)$ and, by condition \ref{regIn}, every element of this set has $s(X, \bz) = s(X, \by_{obs})$. Specifically, set $\bz=\frac{s(X,\boldsymbol{y}_{obs})}{s(X, \boldsymbol{z}^{*})}\bz^{*}$. There are two pieces of this Jacobian: one for the scaling and one for the mapping of the sphere onto $\Pi(\mathcal{A})$. The latter piece is given in equation \eqref{cosine}.
\item Shift  from $\Pi(\mathcal{A})$ to $\mathcal{A}$: $\by=\bz+X\left(\bb(X, \by_{obs})-\bb(X, \bz)\right)$. This shift is along the column space of $X$ to the unique element in $\mathcal{A}$. The Jacobian of this transformation is given by equation \eqref{eq:volume}.
\end{enumerate}
\end{enumerate}

@@

Using the same techniques as in the previous section, 
we fit the normal theory hierarchical model above, a thick-tailed $t$ version with $\nu = 5$ d.f., and two restricted likelihood versions (Huber's and Tukey's) of the model.  For the restricted methods, we condition on robust regression estimates fit separately within each state. We also fit classical robust regression counterparts and a least squares regression separately within each state. 
Additionally, we compare our method to an ABC fit. The ABC version conditions on the Tukey statistics used in our restricted likelihood version. We choose the Tukey version for comparison to ABC since it naturally trims outliers and we expect it to perform the best in this situation. Recall, ABC will approximate the restricted posterior using $\pi(\bth | \rho(T(\by_{obs}), T(\by^{*})) < \epsilon)$. Due to the high-dimension of the parameters and statistics we use the MCMC method called Gibbs ABC developed by \cite{turner2014} to obtain samples from the ABC posterior.
A brief description of this algorithm is as follows with theoretical details provided by  \cite{turner2014}. Let $\by_{j, obs}$ denote the observed data for state $j = 1,\dots, J$.  The shared higher-level parameters are sampled as before since they are, {\em a posteriori} conditionally independent of the data. The state-level parameters $\bth_{j} = (\bbeta_{j}, \sigma_{j}), j = 1, 2, ..., J$ are sampled using Gibbs ABC.  For each iterate of the chain, denote the current state-level parameters and data for state $j$ by $\bth_{j, curr}$ and $\by_{j, curr}$. A single update for the state-level parameters loops over $j$ as follows. Propose $\bth_{j, prop}$ from the prior and then propose new data $\by_{j, prop}$ from the normal model conditional on $\bth_{j, prop}$. The proposed parameters are accepted with Metropolis-Hastings acceptance probability
\begin{align}
\label{eq:mhabc}
\alpha_{mh} = \min\{1, \frac{\phi(\rho(T(\by_{j, prop}), T(\by_{j, obs}))/\delta_{abc})}{\phi(\rho(T(\by_{j, curr}), T(\by_{j, obs}))/\delta_{abc})}\}
\end{align}
with $\phi(\cdot)$ the standard normal pdf and $\delta_{abc}$ a tolerance parameter. Here, we use $\rho(\cdot)$ for the standard Euclidean distance metric to conform to common ABC notation (this is not the $\rho$ of M-estimation). This method makes use of kernel-based ABC \citep{wilkinson2013}. Instead of checking whether the distance between the sampled and observed statistics is strictly below some threshold, this method computes the kernel value of the distance which offers a smoother transition between acceptance and rejection. We use a standard Gaussian kernel for this application. The smaller $\delta_{abc}$, the closer the statistics must be for acceptance.

A few notes on our implementation are warranted. First, we are not restricted to sampling the parameters from their prior. A more general proposal distribution can be applied with standard adjustments to $\alpha_{mh}$ to adjust for the proposal \citep{turner2014}. We tried a few different options, including sampling from the full-conditional distributions, but found proposing from the priors was both the easiest to implement and provided the most satisfactory conversion results for this problem. Additionally, the choice of $\delta_{abc}$ is important and can be different for each state. For this we started with a small value for each state ($\delta_{abc} = 0.01$) and iteratively checked the within-state acceptance rate. After each check, if the rate was below $0.1$ we increased the $\delta_{abc}$ for that state by a factor of $1.2$. These choices were based on some experimentation in order to reach satisfactory convergence in a reasonable number of iterations. To reach satisfactory convergence we had to run the chains for a total of $40,000$ iterations which was $10$-fold more than were needed for the restricted likelihood algorithm. In our experimentation, our method takes only about $1.6$ times as long as ABC per iteration.  We believe that this modest increase in per-iteration computational time is outweighed by apparently better convergence and mixing of the Markov chain.  It is quite possible that better choices for the ABC algorithm could help improve its convergence, but we leave this for further research as computational efficiency is not the main focus of this paper.

@@

The average over states, $\overline{TLM}_b(A)_{\cdot}= \frac{1}{22} \sum_{j =1}^{22} TLM_b(A)_{j}$ for each of the $K$ repetitions is summarized in Figure
\ref{fig:hierTLM} for several trimming fractions using the Student-t as the base model. The points are the average of the $\overline{TLM}_b(A)_{\cdot}$ over the $K$ repetitions with error bars plus/minus one standard deviation over $K$ with larger values representing better predictive performance. As the trimming fraction used for the TLM increases, so does TLM since more outliers are being trimmed. Similar patterns were seen in the individual state level regressions in Section \ref{regModelNW}. Despite being used as the base model to compute TLM, the Student-t doesn't perform well in comparison to the robust regressions. We attribute this to the assumption of heavier tails resulting in smaller log marginal values on average; emphasizing again that the t-model will do well to discount outlying observations but does not provide a natural mechanism for predicting non-outlying data. For each trimming fraction, our restricted likelihood hierarchical models outperform the classical robust regressions fit separately within each state. The hierarchical model also reduces variance in predictions resulting in smaller error bars. 
On the surface, ABC performs quite well in comparison to the restricted likelihood. A table of the mean and standard deviation values for trimming fraction $0.3$ is provided in Table \ref{tab:tlmcompare} where we see that the average TLM for ABC is larger than that for restricted likelihood by $0.03$. Additionally, the standard deviation of TLM for ABC is half the size of that for restricted likelihood. A closer look at the results shows that the difference in average TLM can be attributed entirely to a single state with a relatively small sample size (see next paragraph). 
\begin{table}[H]

>>>>

% commands for for editing
\usepackage{color}
\usepackage{ulem} % \sout
\newcommand{\red}[1]{{\color{red}#1}}
\newcommand{\blue}[1]{{\color{blue}#1}}
\newcommand{\brown}[1]{{\color{brown}#1}}
\newcommand{\green}[1]{{\color{green}#1}}
\newcommand{\magenta}[1]{{\color{magenta}#1}}
\newcommand{\response}[1]{{\color{blue}#1}}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\graphicspath{{figures/}}

@@

\response{Imperfection in the likelihood has also been widely discussed in the classical robustness literature. \cite{hampel1971}, writing on the motivation for studies of robustness, provides a concise description of three mismatches between data and the model that purportedly gives rise to the data:  ``(i) rounding of the observations; (ii) the occurrence of gross errors; (iii) the model itself may only be an approximation to the underlying chance mechanism''.  In a Bayesian setting, the first is easily handled with MCMC methods through the introduction of a latent, unrounded variable into the model.  We do not consider it here.  The second and third are duals.  Misspecification of the model (iii) will often make observations appear to be outliers (ii).  The literature on robust methods is replete with examples described in terms of ``outliers'' where the central problem is model misspecification.  In the sequel, we follow the tradition of referring to cases that are discordant with the stated model as ``outliers'', whether this discordance is due to gross error or a consequence of model misspecification.}

In practice, the imperfections in a proposed likelihood often show themselves through the presence of outliers -- \response{\sout{cases not reflecting the phenomenon under study,}} {\response {whether due to local misspecification of the model or due to gross error}}. There are three main solutions to Bayesian outlier-handling.  The first is to replace the basic sampling density with a mixture model which includes one component for the ``good'' data and a second component for the ``bad'' data.  With this approach, the good component of the sampling density is used for prediction of future good data.  The second approach replaces the
basic sampling density with a thick-tailed density in an attempt to discount outliers, yielding techniques that often provide solid estimates of the center of the distribution but do not easily translate to predictive densities for further good data.  The third approach fits a flexible (typically nonparametric) model to  the data, producing a Bayesian version of a density estimate for both good and bad data.  In recent development, inference is made through the use of robust inference functions \citep{lee2014}.

@@

Restricted likelihoods have also been used as practical approximations to a full likelihood. For example, \cite{pratt1965} appeals to heuristic arguments regarding approximate sufficiency to justify the use of the restricted likelihood of the sample mean and standard deviation. Approximate sufficiency is also appealed to in the use of Approximate Bayesian Computation (ABC), which is related to our method.  
ABC is a collection of posterior approximation methods which has recently experienced success in applications to epidemiology, genetics, and quality control \citep[see, for example,][]{tavare1997, pritchard1999, beaumont2002, marjoram2003, fearnhead2012, drovandi2015}. Interest typically lies in the full data posterior and ABC is used for computational convenience as an approximation.  Consequently, effort is made to choose an approximately sufficient $T(\by)$ and update to the ABC posterior by using the likelihood $L(\bth| \mathcal{B}(\by))$, where $\mathcal{B}(\by)=\{\by^{*}|\rho(T(\by),T(\by^{*})) \leq \epsilon\}$, $\rho$ is a metric, and $\epsilon$ is a tolerance level. This is the likelihood conditioned on the collection of data sets that result in a $T(\cdot)$ within $\epsilon$ of the observed $T(\by)$. %Hence, we can see that, as in our method, conditioning is done on a summary of the complete data. 
With an approximately sufficient $T(\cdot)$ and a small enough $\epsilon$, heuristically  $L(\bth|\mathcal{B}(\by))\approx L(\bth|T(\by))\approx L(\bth|\by)$. Consequently, the ABC posterior approximates the full data posterior and efforts have been made to formalize what is meant by approximate sufficiency \citep[e.g.,][]{joyce2008}. 
\response{
\sout{ABC is related to our method in that the conditioning is on something other than the data $\by$.  However, we specifically seek to condition on an insufficient statistic to guard against misspecification in parts of the likelihood. Additionally, we develop methods where the conditioning is exact (i.e. $\epsilon = 0$).}

Our method can be viewed as ABC with $\epsilon = 0$ and it is natural to compare it to ABC. This paper develops sampling methods for fitting Bayesian linear models conditioning exactly on a set of summary statistics ($\epsilon = 0$), even when the statistics follow a continuous distribution. Traditional ABC sampling methods are flexible and will, in general, apply to a broader class of models. The basic sampling method for ABC is the rejection sampling algorithm \citep{pritchard1999} which proposes a sample $\bth^{*}$ from the prior, then new data $\by^{*}$ from the data-model given $\bth^{*}$. The value $\bth^{*}$ is accepted as a draw from the ABC posterior if $\rho(T(\by),T(\by^{*})) \leq \epsilon$. Acceptance rates of this algorithm can be intolerably low and several extensions have been proposed to improve efficiency \citep[see, for example,][]{beaumont2009, turner2012}. The inefficiency of ABC algorithms is especially problematic in high-dimensional settings since generating high-dimensional statistics that are close to the observed values is difficult. Recently, \cite{turner2014} developed the Gibbs ABC method which improves efficiency in the hierarchical setting by making use of conditional independence of the model to make accept/reject decisions at the individual group-level, effectively reducing the dimension of the problem to the number of parameters within each group. We revisit this approach in our comparisons to ABC in Section \ref{hierRegNW}, finding that, for a modest increase in comptational cost, we obtain an algorithm with better convergence and mixing properties.  We also retain the desired posterior distribution -- the posterior, having conditioned exactly on the summary statistics.
}

@@

\subsection{Illustrative Examples}
\label{illustrations}
Before discussing computational details, the method is applied to two simple examples on well known data sets to demonstrate its effectiveness in situations where outliers are a major concern. The full model in each case fits into the Bayesian linear regression framework discussed in Section \ref{BayesLinMod}. \response{The first is an example (so far as we know) of gross error; the second is an example of model misspecification for a subset of the observations.}  The first example is an analysis of Simon Newcomb's 66 measurements of the passage time of light \citep{stigler1977}; two of which are significant outliers in the lower tail. The full model is a standard location-scale Bayesian model also used in \cite{lee2014}:
The full model in each case fits into the Bayesian linear regression framework discussed in Section \ref{BayesLinMod}. The first example is an analysis of Simon Newcomb's 66 measurements of the passage time of light \citep{stigler1977}; two of which are significant outliers in the lower tail. The full model is a standard location-scale Bayesian model also used in \cite{lee2014}:\begin{equation}
\beta\sim N(23.6, 2.04^{2}),\  \sigma^{2}\sim IG(5, 10), \ y_{i}\iid N (\beta, \sigma^{2}), i=1,2,\dots, n=66,
\end{equation}
where $y_{i}$ denotes the $i^{th}$ (recorded) measurement of the passage time of light. $\beta$ is interpreted as the passage time of light with the deviations $y_i - \beta$ representing measurement error.
Four versions of the restricted likelihood are fit with conditioning statistics: 1) Huber's M-estimator for location with Huber's `proposal 2'  for scale 2)  Tukey's M-estimator for location with Huber's `proposal 2'  for scale 3) LMS (least median squares) for location with associated estimator of scale and 4) LTS (least trimmed squares)  for location with associated estimator of scale. Details of these estimators can be found in many places, including \citep{huber2009}. We return to the two M-estimators throughout this paper as we have found them to offer good default choices for practitioners dealing with outliers. A short review of these estimators is provided in the Supplementary Material.  The tuning parameters for the M-estimators are chosen to achieve $95\%$ efficiency under normality \citep{huber2009} and, for comparability, roughly $5\%$ of the residuals are trimmed for LTS.  Two additional approaches to outlier handling are considered: 1) the normal distribution is replaced with a t-distribution and, 2) the normal distribution is replaced with a mixture of two normals. The t-model assumes $y_{i}\iid t_{\nu} (\beta, \sigma^{2})$ with $\nu=5$. The prior on $\sigma^{2}$ is $IG(5, \frac{\nu-2}{\nu}10)$ and ensures that the prior on the variance is the same as the other models. The mixture takes the form: $y_{i}\iid pN (\beta, \sigma^{2}) + (1-p)N(\beta, 10\sigma^{2})$ with the prior $p \sim \text{beta}(20,1)$ on the probability of belonging to the `good' component.

The posterior of $\beta$ under each model appears in Figure \ref{fig:newcomb_post}.  The posteriors group into two batches.  The normal model and restricted likelihood with LMS do not discount the outliers and have posteriors centered at low values of $\beta$.  These posteriors are also quite diffuse.  In contrast, the t-model, mixture model, and the other restricted likelihood methods discount the outliers and have posteriors centered at higher values.  There is modest variation among these centers.  Posteriors in this second group have less dispersion than those in the first group.   
\begin{figure}[t]
\centering
{\includegraphics[width = 4in]{speed_of_light_beta.png}}
{\includegraphics[width = 4in]{speed_of_light_predictive.png}}
\caption{Results from the analysis of the speed of light data. Top: Posterior distributions of $\beta$ under each model. Bottom: Log posterior predictive distributions under each model. The differences in the tails are emphasized in the bottom plot. The horizontal axis is strategically labeled to help compare the centers of the distributions in each of the plots.}
\label{fig:newcomb_post}
\end{figure}
The pattern for predictive distributions differs (see bottom plot in Figure \ref{fig:newcomb_post}).  The normal and t-models have widely dispersed predictive distributions.  The other predictive distributions show much greater concentration.  The restricted likelihood fits based on M-estimators (Tukey's and Huber's) are centered appropriately and are concentrated. The restricted likelihood based on LTS and the mixture model results are also centered appropriately, but comparatively less concentrated. The LMS predictive is concentrated, but it is poorly centered.

@@

The data and  $95\%$ credible bands for the posterior predictive distribution under each model are displayed in Figure \ref{fig:calls_predictive}. The normal model fit to all cases results in a very wide posterior predictive distribution due to an inflated estimate of the variance. The t-model provides a similar predictive distribution.  The pocket of outliers from 1963 to 1970 overwhelms the natural robustness of the model and leads to wide prediction bands.  The outliers, falling toward the end of the time period, lead to a relatively high slope for the regression.  In contrast, the normal theory model fit to only the good data results in a smaller slope and narrower prediction bands.  The predictive distribution under the restricted likelihood approach is much more precise and is close to that of the normal theory fit to the non-outlying cases. The two component mixture model provides similar results, where the predictive distribution is formulated using only the good component. For these data, the large outliers are easily identified as following a distinct regression, leaving the primary component of the mixture for non-outlying data.  In a more complex situation where the outlier generating mechanism is transient (i.e., ever changing and more complex than for these data), modeling the outliers is more difficult. As in classical robust estimation, the restricted likelihood approach avoids explicitly modeling the outliers. 
\begin{figure}[t]
\centering
{\includegraphics[width = 4in]{calls_predictive.png}}
\caption{Pointwise posterior predictive intervals of log(calls) under the normal theory model fit to the non-outliers, the restricted likelihood model with Tukey's M-estimator for the slope and intercept with Huber's `proposal 2'  for scale, and a heavy-tailed t-distribution model. The first three data points were used to specify the prior with each model using the remaining 21 for fitting. The normal theory model was also fit after removing observations 14-20 (years 1963 - 1970).}
\label{fig:calls_predictive}
\end{figure}

@@

Given the one-to-one and onto mapping $h: \mathbb{S} \rightarrow \mathcal{A}$, the general proposal strategy is summarized as follows:
\begin{enumerate}
\item Sample $\bz^*$ from a distribution with known density \response{whose support is the entirety of} $\mathbb{S}$.
\item Set $\by = h(\bz^*)$ and calculate the Jacobian of this transformation in two steps.
\begin{enumerate}
\item Scale from $\mathbb{S}$ to the set $\Pi(\mathcal{A}):= \{\bz\in \mathbb{R}^n |\ \exists\ \by\in \mathcal{A}\ s.t.\ \bz=Q \by \}$ with $Q = I - XX^{\top}$. \footnote{We have used condition \ref{fullRank} to assume  without loss of generality  that the columns of $X$ form an orthonormal basis for $\mc{C}(X)$ (i.e., $X^\top X=I$).} $\Pi(\mathcal{A})$ is the projection of $\mathcal{A}$ onto $\mathcal{C}^{\perp}(X)$ and, by condition \ref{regIn}, every element of this set has $s(X, \bz) = s(X, \by_{obs})$. Specifically, set $\bz=\frac{s(X,\boldsymbol{y}_{obs})}{s(X, \boldsymbol{z}^{*})}\bz^{*}$. There are two pieces of this Jacobian: one for the scaling and one for the mapping of the sphere onto $\Pi(\mathcal{A})$. The latter piece is given in equation \eqref{cosine}.
\item Shift  from $\Pi(\mathcal{A})$ to $\mathcal{A}$: $\by=\bz+X\left(\bb(X, \by_{obs})-\bb(X, \bz)\right)$. This shift is along the column space of $X$ to the unique element in $\mathcal{A}$. The Jacobian of this transformation is given by equation \eqref{eq:volume}.
\end{enumerate}
\end{enumerate}

@@

Using the same techniques as in the previous section, 
we fit the normal theory hierarchical model above, a thick-tailed $t$ version with $\nu = 5$ d.f., and two restricted likelihood versions (Huber's and Tukey's) of the model.  For the restricted methods, we condition on robust regression estimates fit separately within each state. We also fit classical robust regression counterparts and a least squares regression separately within each state. 
\response{
Additionally, we compare our method to an ABC fit. The ABC version conditions on the Tukey statistics used in our restricted likelihood version. We choose the Tukey version for comparison to ABC since it naturally trims outliers and we expect it to perform the best in this situation. Recall, ABC will approximate the restricted posterior using $\pi(\bth | \rho(T(\by_{obs}), T(\by^{*})) < \epsilon)$. Due to the high-dimension of the parameters and statistics we use the MCMC method called Gibbs ABC developed by \cite{turner2014} to obtain samples from the ABC posterior.
A brief description of this algorithm is as follows with theoretical details provided by  \cite{turner2014}. Let $\by_{j, obs}$ denote the observed data for state $j = 1,\dots, J$.  The shared higher-level parameters are sampled as before since they are, {\em a posteriori} conditionally independent of the data. The state-level parameters $\bth_{j} = (\bbeta_{j}, \sigma_{j}), j = 1, 2, ..., J$ are sampled using Gibbs ABC.  For each iterate of the chain, denote the current state-level parameters and data for state $j$ by $\bth_{j, curr}$ and $\by_{j, curr}$. A single update for the state-level parameters loops over $j$ as follows. Propose $\bth_{j, prop}$ from the prior and then propose new data $\by_{j, prop}$ from the normal model conditional on $\bth_{j, prop}$. The proposed parameters are accepted with Metroplis-Hastings acceptance probability
\begin{align}
\label{eq:mhabc}
\alpha_{mh} = \min\{1, \frac{\phi(\rho(T(\by_{j, prop}), T(\by_{j, obs}))/\delta_{abc})}{\phi(\rho(T(\by_{j, curr}), T(\by_{j, obs}))/\delta_{abc})}\}
\end{align}
with $\phi(\cdot)$ the standard normal pdf and $\delta_{abc}$ a tolerance parameter. Here, we use $\rho(\cdot)$ for the standard Euclidean distance metric to conform to common ABC notation (this is not the $\rho$ of M-estimation). This method makes use of kernel-based ABC \citep{wilkinson2013}. Instead of checking whether the distance between the sampled and observed statistics is strictly below some threshold, this method computes the kernel value of the distance which offers a smoother transition between acceptance and rejection. We use a standard Gaussian kernel for this application. The smaller $\delta_{abc}$, the closer the statistics must be for acceptance.

A few notes on our implementation are warranted. First, we are not restricted to sampling the parameters from their prior. A more general proposal distribution can be applied with standard adjustments to $\alpha_{mh}$ to adjust for the proposal \citep{turner2014}. We tried a few different options, including sampling from the full-conditional distributions, but found proposing from the priors was both the easiest to implement and provided the most satisfactory conversion results for this problem. Additionally, the choice of $\delta_{abc}$ is important and can be different for each state. For this we started with a small value for each state ($\delta_{abc} = 0.01$) and iteratively checked the within-state acceptance rate. After each check, if the rate was below $0.1$ we increased the $\delta_{abc}$ for that state by a factor of $1.2$. These choices were based on some experimentation in order to reach satisfactory convergence in a reasonable number of iterations. To reach satisfactory convergence we had to run the chains for a total of $40,000$ iterations which was $10$-fold more than were needed for the restricted likelihood algorithm. In our experimentation, our method takes only about $1.6$ times as long as ABC per iteration.  We believe that this modest increase in per-iteration computational time is outweighed by apparently better convergence and mixing of the Markov chain.  It is quite possible that better choices for the ABC algorithm could help improve its convergence, but we leave this for further research as computational efficiency is not the main focus of this paper.}

@@

The average over states, $\overline{TLM}_b(A)_{\cdot}= \frac{1}{22} \sum_{j =1}^{22} TLM_b(A)_{j}$ for each of the $K$ repetitions is summarized in Figure
\ref{fig:hierTLM} for several trimming fractions using the Student-t as the base model. The points are the average of the $\overline{TLM}_b(A)_{\cdot}$ over the $K$ repetitions with error bars plus/minus one standard deviation over $K$ with larger values representing better predictive performance. As the trimming fraction used for the TLM increases, so does TLM since more outliers are being trimmed. Similar patterns were seen in the individual state level regressions in Section \ref{regModelNW}. Despite being used as the base model to compute TLM, the Student-t doesn't perform well in comparison to the robust regressions. We attribute this to the assumption of heavier tails resulting in smaller log marginal values on average; emphasizing again that the t-model will do well to discount outlying observations but does not provide a natural mechanism for predicting non-outlying data. For each trimming fraction, our restricted likelihood hierarchical models outperform the classical robust regressions fit separately within each state. The hierarchical model also reduces variance in predictions resulting in smaller error bars. 
\response{
On the surface, ABC performs quite well in comparison to the restricted likelihood. A table of the mean and standard deviation values for trimming fraction $0.3$ is provided in Table \ref{tab:tlmcompare} where we see that the average TLM for ABC is larger than that for restricted likelihood by $0.03$. Additionally, the standard deviation of TLM for ABC is half the size of that for restricted likelihood. A closer look at the results shows that the difference in average TLM can be attributed entirely to a single state with a relatively small sample size (see next paragraph). }
\begin{table}[H]
